### 2024-06-18 README.Rmd
Moving on to what AI does to everyone and everything

> Mission Statement: Our organization seeks to level up on all matters related.  
> -- An Organization Run by its Marketing Department

@@

In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 75 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone. As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems. For instance, @mollick_instructors_2024 note that "Large Language Models are trained in ways that may violate copyright, and often rely on the efforts of low-wage workers in precarious conditions for part of their training. Models are trained on biased data and can produce either subtly or overtly biased results. And because these biases seem to come from an objective machine, they can be especially pernicious. Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems". This disclaimer, consuming fully one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highlights the important role of institutions to provide the moral and ethical clarity that may elude individual instructors working under desperately under-resourced conditions, or simply willing to win a race to the bottom.].

@@

@@

Neural networks have been used for decades to do data analysis. There is nothing new here. But what is being promised is that we will be able to do data analysis without having to learn how to do data analysis. That somehow a chatbot will do it for us. The idea that a chatbot could be used to do data analysis betrays a profound misunderstanding of what chatbots do, and almost absolute ignorance of what data analysis entails. Ask a chatbot to do some arithmetic: it will produce something that is right on average. But average is not always right. And seeing how AIs are now being trained on the dregs of the internet (see below) it should not surprise anyone when an AI Overview responds to the prompt "2 + 2" with "potato".

@@

An error like this is easy to spot. But we seldom use computers to do calculations that can be done by hand, and checking that an AI-generated response is not wrong is likely to be as time consuming as not using an AI to do data analysis in the first place [see @yuan_how_2023].

@@

@@

The preceding section illustrates how numerous tasks that AIs can notionally do in higher education are in fact poor replacements for a human-centered education--and that is even before considering the moral, ethical, and possibly legal compromises that the use of corporate-developed AIs demand (non-corporate developed AIs can be ignored: as @vandehei_behind_2024 remind us, only a handful of corporations have the resources to make AIs give the impression of competence). But what about the things that GenAIs can do _to_ students?

@@

But let us assume for a moment that using AIs in their current form does becomes an in-demand tool in the short term. What is the expected longevity of skills like prompt "engineering"?

Apart from a few methodological developments (like transformers), improvement in the performance of AIs (say, from ChatGPT-2 to the latest version), mostly are the result of using bigger datasets. This has led to a rush, as developers try to improve their models by training them on more and more data^[A rush that feels at least as frenetic as any gold rush in the past. And, while the gold rushes of the 19th century were relatively democratic due to their low entry costs--both in terms of immigration and tools--the current data rush has enormous entry costs and is largely dominated by venture capital (that is, equity investments that support pre-launch, launch, and early stages of a business) and established large corporations.]. But, despite the large number of text items and images available in digital form, there is still only a finite amount of data to be had.

@@

As the fountain of human-generated content, generously shared by countless people over a period of decades, becomes exhausted by a handful of corporations developing AIs, there has been a turn towards what might be the last juicy nuggets of information left on the internet.

OpenAI, as a leader in data grabbing, turned to Stack Overflow for more data [@stack_overflow_stack_2024]. Stack Overflow [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges, and one of the most popular websites in the world". Said community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) (to say the least) with the way Stack Overflow dealt with OpenAI, and has come to believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution, something that AIs notoriously _do not do_. Some members of the community protested the deal with OpenAI by trying to withdraw or change their posts, in what Ars Technica unhelpfully termed [sabotage](https://arstechnica.com/information-technology/2024/05/stack-overflow-users-sabotage-their-posts-after-openai-deal/). It is an open question how long Stack Overflow will remain a community-based space, especially after it started treating its members as unwitting, and even unwilling providers of data for OpenAI.

@@

Mind you, these, and numerous other hapless results^[Including AI overviews with misinformation about depression [@tenbarge_glue_2024].] are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024]. To what extent can more money improve GenAIs? Probably not by much, since a key constraint is the amount of data available for training the deep neural networks behind the technology [@villalobos_position_2024]. And dismal as this is, the outlook may be even worse than simply a frantic race after more human-generated data^[A disturbing idea in and of itself, given the appetite for risky behaviors by corporate actors, and the potential for abuse of nominally private and confidential information. It would be too naive to expose even a morsel of juicy data to transmission over the internet, and expect Google et al., not to see it as fair game.]. As GenAIs become more widely used, they can produce data at a pace that humans cannot match, much of which is already finding its way into the internet in the form of text and images--and not only through clickbait websites^[A blog post on a website called _Ancient Rome_ by a certain Ellen Hunter (whose profile picture is likely the output of a generative AI and who does not seem to exist elsewhere) discusses the topic of windows in classical Rome and Greece [@hunter_did_2023]. Under subheading 4.1 (_What Existed Before Windows?_), the blog explains that "MS-DOS was a disk operating system that was developed by Microsoft. It was the predecessor to the Windows operating system and was used on IBM PC compatible computers. MS-DOS was first released in 1981 and was the most popular operating system for personal computers during the 1980s". Clearly, the text is the output of a chatbot, and was posted to the internet without regard to its accuracy; but why waste time on even a cursory check, when the objective of the blog is not to inform, but to attract clicks for monetizing ads?] and disinformation [@ruffo_studying_2023], but also in scientific writing, supposedly the most thoroughly vetted form of writing in the world [@maiberg_ai_2024]^[And let us not fool ourselves with the idea that this will be limited to smaller, scammy paper mills, when examples begin to crop up in publications by IEEE, Elsevier, and Springer, as evidenced by the telltale use of "tortured phrases" [@cabanac_tortured_2021] and chatbot stock expressions like "[a]s of my last knowledge update" [e.g., @swain_deep_2023; @al-worafi_pharmacy_2023; @sumit_dynamic_2024; @zhang_retracted_2024].].

@@

The implication should be clear: training models with the output of models is a recipe for less varied outputs. We may be witnessing _Peak ChatGPT_, and future versions of AIs, instead of improving, may begin to regress to a bland version of their own previous outputs. And this is before we even consider the possibility of AIs being deliberately "poisoned". As backlash against a data rush that is perceived as exploitative and abusive grows, new tools are being developed to generate data that are manipulated in ways that are too subtle to be detectable by a human, but that are deleterious to the training process of a GenAI [@salman_raising_2023]. One such tool is @shan_nightshade_2024's _Nightshade_, an algorithm that works to corrupt prompts used in text-to-image generation, and that notably bleeds through to poison related concepts. Another tool is @xu_shadowcast_2024's _Shadowcast_, an algorithm that poisons samples to trick an AI into misidentifying class labels, and that also can perform persuasion attacks, leveraging image-to-text generation capabilities to craft nonsensical narratives.

@@

@@

But AIs must do something for someone, otherwise why would they be everywhere? What do AIs do for the likes of OpenAI, Microsoft, and Google? So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money. Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.], much of it poured into autonomous vehicles and related mobility tech, followed by healthcare, biotechnologies and drugs, with business processes and support services in third place [@tricot_venture_2021].

@@

But the industry has not necessarily made money for investors yet. For example, @berber_peter_2024 reports that the AI industry spent \$50 billion (USD) in 2023 on chips alone, while bringing in only $3 billion in revenue. This has not prevented start-ups from pivoting to GenAI, in a shift  of focus to this much hyped technology from whatever they were doing in the first place^[As if any other signs of the scammy environment in which GenAI tech thrives were needed, @berber_peter_2024 also reports how Cognition, a startup that pivoted to AI from cryptocurrency, was receiving billionaire infusions of speculative money.].

@sriram_ai_2023

By some accounts   @amdur_venture_2023

Distractions:
@vinsel_dont_2023 @siegel_ai_2023

Deskilling leads to...

So far, not money. But since corporations are not in the development of AI for charitable motives, at some point they will want to make money out of them. Who will pay? And how much will they need to pay? Or have to pay, if their skills are inexorable dependent on this one technology?

To read:

Charlie Pierce [writes](https://www.antipope.org/charlie/blog-static/2024/04/the-radiant-future-of-1995.html#more) that "the industry is running at a loss on power consumption alone, never mind labour costs (which are quite high: those generative LLMs require extensive human curation of the input data they require for training)".

"The common feature of these bubbles is a shitload of hype and promises from hucksters who fail to deliver a viable product but suck up as much investment capital as they can. A handful of them survive: from dot-com 1.0, the stand-outs are Amazon and Google (Facebook, Twitter, Reddit, et al came along much later—social media was a later, smaller bubble). Other survivors include Paypal, eBay, and Doubleclick (the latter being merged with Google to form a monstrous global advertising monopoly). The survivors tend to leave behind infrastructure: the failures leave behind t-shirts, second hand Aeron chairs, and motivational posters."

https://futurism.com/the-byte/google-bard-ai-ready NOT READY, SO WHY RELEASE IT?

@@

```{r puzzle-web-search, out.width="50%", fig.cap="\\label{fig:puzzle-web-search}Figure 15. A search using Duck Duck Go"}
ggdraw() + 
  draw_image(image = "images/logical-puzzle-web-search.png")
```

```{r puzzle-chatgpt, out.width="50%", fig.cap="\\label{fig:puzzle-chatgpt}Figure 16. A conversation with ChatGPT-4"}
puzzle_plot <- ggplot() +
  labs(caption = "https://assemblag.es/@hugo/112450401657084750") +
  theme_void()

>>>>

> _Mission Statement: Our organization seeks to level up on all matters related._  
> -- An Organization Run by its Marketing Department

@@

In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 80 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone. As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems. For instance, @mollick_instructors_2024 note that "Large Language Models are trained in ways that may violate copyright, and often rely on the efforts of low-wage workers in precarious conditions for part of their training. Models are trained on biased data and can produce either subtly or overtly biased results. And because these biases seem to come from an objective machine, they can be especially pernicious. Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems". This disclaimer, consuming fully one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highlights the important role of institutions to provide the moral and ethical clarity that may elude individual instructors working under desperately under-resourced conditions, or simply willing to win a race to the bottom.].

@@

> _ChatGPT doesn't give you information. It gives you information-shaped sentences._
> --Neil Gaiman

@@

Neural networks have been used for decades to do data analysis. There is nothing new here. But what is being promised is that we will be able to do data analysis without having to learn how to do data analysis. That somehow a chatbot will do it for us. The idea that a chatbot could be used to do data analysis betrays a profound misunderstanding of what chatbots do, and almost absolute ignorance of what data analysis entails. Ask a chatbot to do some arithmetic: it will produce something that is right on average. But average is not always right. And seeing how AIs are now being trained on the dregs of the internet, it should surprise approximately no one when an AI Overview responds to the prompt "2 + 2" with "potato".

@@

An error like this is easy to spot. But we seldom use computers to do calculations that can be done by hand, and checking that an AI-generated response is not wrong is likely to be as time consuming as not using an AI to do quantitative data analysis in the first place [see @yuan_how_2023].

@@

But what about a task that involves logic instead of arithmetic? LLMs are good for it only if the goat knows how to swim.

```{r puzzle-web-search, eval=FALSE, out.width="50%", fig.cap="\\label{fig:puzzle-web-search}Figure 15. A search using Duck Duck Go"}
ggdraw() + 
  draw_image(image = "images/logical-puzzle-web-search.png")
```

```{r puzzle-chatgpt, out.width="50%", fig.cap="\\label{fig:puzzle-chatgpt}Figure 16. ChatGPT-4 makes the goat swim"}
puzzle_plot <- ggplot() +
  labs(caption = "https://assemblag.es/@hugo/112450401657084750") +
  theme_void()

As research by @nezhurina_alice_2024 demonstrates, relatively simple natural language puzzles that are easily solved by humans, often result in confidently expressed nonsense when used to prompt an AI. Wrong answers, furthermore, are frequently backed up by confabulations, that is, reasoning-like or plausible sounding explanations that are illogical. If anything, mathematics should be easy: after all, computers have been doing mathematics since at least the [Antikythera mechanism](https://en.wikipedia.org/wiki/Antikythera_mechanism) (ca. 178 BC), which preceded other mechanical machines by centuries--including Babbage's [Analytical Engine](https://en.wikipedia.org/wiki/Analytical_engine), the machine that [Ada Lovelace](https://en.wikipedia.org/wiki/Ada_Lovelace) used to code an algorithm to write Bernoulli numbers. Electronic computers have been doing mathematics reliably for decades now. Computers can do mathematics well because they are implemented using a crisp, unambiguous language. Natural language is full of nuances that are not easily captured using the medium of text alone. The stochastic connections between words that LLMs produce are simply probabilistic word salads that only sound plausible thanks to the billions of data points used to train them and the billions of parameters that the models use. But if the objective of qualitative research is to identify interesting aspects of the human experience, it is certainly not the case that they will be found in a regression to the mean.

@@

The preceding section illustrates how numerous tasks that AIs can notionally do in higher education are in fact poor replacements for a human-centered education--and that is even before considering the moral, ethical, and possibly legal compromises that the use of corporate-developed AIs demand (non-corporate developed AIs can be ignored: as @vandehei_behind_2024 remind us, only a handful of corporations have the resources needed to create the illusion of compentence in an AI).

But what about the things that GenAIs can do _to_ students?

@@

However, let us assume for a moment that using AIs in their current form does becomes an in-demand tool in the short term. What is the expected longevity of skills like prompt "engineering"?

Apart from a few methodological developments (like transformers), improvement in the performance of AIs (say, from ChatGPT-2 to the latest version), mostly are the result of using bigger datasets. This has led to a rush, as developers try to improve their models by training them on more and more data^[A rush that feels at least as frenetic as any gold rush in the past. And, while the gold rushes of the 19th century were relatively democratic due to their low entry costs--both in terms of immigration and tools--the current data rush has enormous entry costs and is largely dominated by large corporations and venture capital (that is, equity investments that support pre-launch, launch, and early stages of a business).]. But, despite the large number of text items and images available in digital form, there is still only a finite amount of data to be had.

@@

As the fountain of human-generated content--generously (one might even say naively) shared by untold numbers of people over a period of decades--becomes exhausted by a handful of corporations developing AIs, there has been a turn towards what might be the last juicy nuggets of information left on the internet.

OpenAI, a global leader in data grabbing, turned to Stack Overflow for more data [@stack_overflow_stack_2024]. Stack Overflow [bills itself](https://stackoverflow.com) as "[a] community-based space to find and contribute answers to technical challenges, and one of the most popular websites in the world". Said community was [unhappy](https://favtutor.com/articles/stack-overflow-community-not-happy-openai/) (to say the least) with the way Stack Overflow dealt with OpenAI, and has come to believe that the terms were abusive: while Stack Overflow's terms of service state that posters cannot revoke permission to use, the site also operates under a Creative Commons license that requires attribution, something that AIs notoriously _do not do_. Some members of the community protested the deal with OpenAI by trying to withdraw or change their posts, in what Ars Technica unhelpfully termed [sabotage](https://arstechnica.com/information-technology/2024/05/stack-overflow-users-sabotage-their-posts-after-openai-deal/). It is an open question how long Stack Overflow will remain a community-based space, especially after it started treating its members as unwitting, and even unwilling providers of data for OpenAI [@grimm_stack_2024].

@@

Mind you, these, and numerous other hapless results^[Including AI overviews with misinformation about depression [@tenbarge_glue_2024].] are produced by an AI that Google has already spent tens of billions of dollars developing [@vynck_after_2024]. To what extent can more money improve GenAIs? Probably not by much, since a key constraint is the amount of data available for training the deep neural networks behind the technology [@villalobos_position_2024]. And dismal as this is, the outlook may be even worse than simply a frantic race after more human-generated data^[A disturbing idea in and of itself, given the appetite for risky behaviors by corporate actors, and the potential for abuse of nominally private and confidential information. It would be too naive to expose even a morsel of juicy data to transmission over the internet, and expect Google et al., not to see it as fair game.]. As GenAIs become more widely used, they can produce data at a pace that humans cannot match, much of which is already finding its way into the internet in the form of text and images--and not only through clickbait websites^[A blog post on a website called _Ancient Rome_ by a certain Ellen Hunter (whose profile picture is likely the output of a generative AI and who does not seem to exist elsewhere) discusses the topic of windows in classical Rome and Greece [@hunter_did_2023]. Under subheading 4.1 (_What Existed Before Windows?_), the blog explains that "MS-DOS was a disk operating system that was developed by Microsoft. It was the predecessor to the Windows operating system and was used on IBM PC compatible computers. MS-DOS was first released in 1981 and was the most popular operating system for personal computers during the 1980s". Clearly, the text is the output of a chatbot, and was posted to the internet without regard to its accuracy; but why waste time on even a cursory check, when the objective of the blog is not to inform, but to attract clicks for monetizing ads?] and disinformation [@ruffo_studying_2023], but also in scientific writing, supposedly the most thoroughly vetted form of writing in the world [@maiberg_ai_2024]^[And let us not fool ourselves with the idea that this will be limited to smaller, scammy paper mills, when examples begin to crop up in publications by IEEE, Elsevier, and Springer, as evidenced by the telltale use of "tortured phrases" [@cabanac_tortured_2021] and chatbot stock expressions like "[a]s of my last knowledge update...", "[a]s of my knowledge cutoff...", "[c]ertainly! Here are...", and "[r]egenerate response" [e.g., @al-worafi_pharmacy_2023; @farag_exploring_2023; @li_revitalizing_2023; @shafiq_escrao_2023; @swain_deep_2023; @alsaif_nd2o3-_2024; @gatell_impact_2024; @remzan_ensemble_2024; @sumit_dynamic_2024; @zhang_retracted_2024]. One particularly egregious example straight out includes the phrase "[p]lease note that as an AI language model, I am unable to generate specific tables or conduct tests, so the actual results should be included in the table" [@yang_utilization_2023].].

@@

The implication should be clear: training models with the output of models is a recipe for increasingly bland outputs. We may be witnessing _Peak ChatGPT_, and future versions of AIs, instead of improving, may begin to regress to duller versions of their own previous outputs. And this is before we even consider the possibility of AIs being deliberately "poisoned". As backlash against a data rush that is perceived as exploitative and abusive grows, new tools are being developed to generate data that are manipulated in ways that are too subtle to be detectable by a human, but that are deleterious to the training process of a GenAI [@salman_raising_2023]. One such tool is @shan_nightshade_2024's _Nightshade_, an algorithm that works to corrupt prompts used in text-to-image generation, and that notably bleeds through to poison related concepts. Another tool is @xu_shadowcast_2024's _Shadowcast_, an algorithm that poisons samples to trick an AI into misidentifying class labels, and that also can perform persuasion attacks, leveraging image-to-text generation capabilities to craft nonsensical narratives.

@@

> _Teach a person to use ChatGPT, and they will be in thrall to a corporation for as long as that is their skill. Teach them to think, and they will be free for life._  
> -- Anonymous

@@

But AIs must do something for someone, otherwise why would they be everywhere? What do AIs do for the likes of OpenAI, Microsoft, and Google? So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities, such as the resources needed to have a University Committee to advise on AI^[Some observers have even wondered whether the AI fever is not actually becoming detrimental to innovation [@siegel_ai_2023; @vinsel_dont_2023].]. Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.], much of it poured into autonomous vehicles and related mobility tech, followed by healthcare, biotechnologies and drugs, with business processes and support services in third place [@tricot_venture_2021].

@@

Alas, the industry has not necessarily made money for investors yet. For example, @berber_peter_2024 reports that the AI industry spent \$50 billion (USD) in 2023 on chips alone, while bringing in only $3 billion in revenue. @patel_inference_2023 estimated that incorporating ChatGPT-like LLMs into search would lead to a loss of \$30 billion (USD) in profit for Google. At a guess, the industry is running at a loss: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) for OpenAI to begin losing money on that one customer. And a large number of users are not even paying subscribers: based on revenue reports [see @perez_chatgpts_2024] there were at most 210,000 subscribers to the "Plus" plan in 2024, out of a user base estimated at tens of millions [@tong_exclusive_2023]. The profitability of AI would look even worse if not for the fact that the cost of labor is suppressed by outsourcing the most intensive, disturbing, degrading, and damaging aspects of training AIs to the Global South [@nix_analysis_2024]. The fact that the industry is not profitable has not prevented start-ups from pivoting to GenAI, in a shift  of focus to this much hyped technology from whatever they were doing in the first place^[As if any other signs  were needed of the scammy environment in which GenAI tech thrives, @berber_peter_2024 also reports how Cognition, a startup backed by Peter Thiel, pivoted to AI from cryptocurrency, after cryptocurrency went sour with the frauds at FTX.].

Venture capital is not known for its charitable aims, though, and at some point investors will want their money back with a heap of profit on top [@sriram_ai_2023]. Who will pay to make investors whole? And how much will those who end up paying need to pay? Or, in the worst case, _have_ to pay, if their skills are inexorable tied to this one technology? Deskilling, both of individuals and of organizations, appears to be one of the few pathways to AI profitability: once people have failed to learn a skill, and therefore become dependent on the tool, it becomes possible for the maker of the tool to turn into a digital rentier [@sadowski_internet_2020]. From this perspective, what is on offer by the AI industry is SaaS--Skill as a Service^[Universities, instead of being innovators, have increasingly become renters in the land of the platform economy [@komljenovic_rise_2021].].

@@
